---
title: "TCGA-BRCA RNA-Seq Data Mining: Preprocessing and Machine Learning"
author: "Ebenezer Oppong Gyamfi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(readr)

#LOADING TCGA DATA
tcga_data <- read_tsv("biodata/BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt")

head(tcga_data)
```

# 1. Introduction

This report documents the complete data-mining workflow applied to the
TCGA Breast Invasive Carcinoma (BRCA) RNA-Seq Level 3 mRNAseq dataset
(RSEM output, Broad GDAC Firehose pipeline). The dataset encompasses
**878 samples** and an initial **20,503 gene identifiers**. The pipeline
covers: data structuring, noise removal, missing-value handling,
descriptive statistics, visualization, transformation, dimensionality
reduction, Principal Component Analysis (PCA), and sample proximity
measures — following the framework from the Data Preprocessing, Basic
Statistics, and Proximity Measures sessions. Every method is chosen
deliberately and explained in place.

# 2. Data Structuring

The raw TCGA export stores three value columns per sample (raw counts,
median-length-normalized counts, RPKM) under repeated sample barcodes,
and encodes each gene as `SYMBOL|ENTREZ_ID`. R appends `...N` suffixes
to deduplicate the repeated column headers, and the true value-type
labels live in row 1 of the data rather than in the header row. Three
structural operations were therefore required before analysis.

```{r structure-data}
# Step 1: Recover true barcodes by stripping R's auto-added "...N" suffix
barcodes     <- sub("\\.\\.\\.\\d+$", "", colnames(tcga_data))
barcodes[1]  <- "gene"
value_types  <- as.character(tcga_data[1, ])
value_types[1] <- ""
new_names    <- ifelse(value_types == "",
                       barcodes,
                       paste(barcodes, value_types, sep = "_"))
colnames(tcga_data) <- new_names

# Drop the embedded header row
expr_raw <- tcga_data[-1, ]

# Step 2: Split "SYMBOL|ENTREZ" gene identifier
expr_raw <- expr_raw %>%
  separate(gene, into = c("symbol", "entrez_id"), sep = "\\|", remove = FALSE)

# Step 3: Remove unmapped genes (symbol == "?")
n_before <- nrow(expr_raw)
expr_raw  <- expr_raw %>% filter(symbol != "?")
cat("Unmapped ('?') genes removed:", n_before - nrow(expr_raw),
    "out of", n_before, "\n")
```

```{r select-counts}
# Work with raw_counts only
counts_df  <- expr_raw %>% select(gene, symbol, entrez_id,
                                   ends_with("_raw_counts"))
count_cols <- grep("_raw_counts$", colnames(counts_df), value = TRUE)
counts_df[count_cols] <- lapply(counts_df[count_cols], as.numeric)

# Step 4: Collapse duplicate gene symbols — keep most-expressed row per symbol
n_dup <- sum(duplicated(counts_df$symbol) |
             duplicated(counts_df$symbol, fromLast = TRUE))
cat("Rows involved in duplicate gene symbols:", n_dup, "\n")

counts_df <- counts_df %>%
  mutate(row_total = rowSums(across(all_of(count_cols)), na.rm = TRUE)) %>%
  group_by(symbol) %>%
  slice_max(row_total, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-row_total)

# Build the numeric expression matrix
expr_mat <- as.matrix(counts_df[, count_cols])
rownames(expr_mat) <- counts_df$symbol
colnames(expr_mat) <- sub("_raw_counts$", "", count_cols)

cat("Matrix after structuring:", nrow(expr_mat), "genes x",
    ncol(expr_mat), "samples\n")
```

```{r sample-metadata}
# Derive sample group from TCGA barcode (positions 14-15 = sample-type code)
# 01-09 = tumor, 06 = metastatic, 10-19 = normal
sample_type_code <- substr(colnames(expr_mat), 14, 15)
sample_group <- case_when(
  as.numeric(sample_type_code) == 6              ~ "Metastatic",
  as.numeric(sample_type_code) < 10             ~ "Tumor",
  TRUE                                           ~ "Normal"
)
sample_meta <- data.frame(
  sample = colnames(expr_mat),
  type_code = sample_type_code,
  group = sample_group
)
table(sample_meta$group)
```

```{r sample-composition-plot, fig.width=5, fig.height=4}
# Figure 1: Sample composition
comp_df <- as.data.frame(table(Group = sample_meta$group))
ggplot(comp_df, aes(x = Group, y = Freq, fill = Group)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = Freq), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c("Tumor"      = "#B22222",
                                "Normal"     = "#2E7D6E",
                                "Metastatic" = "#6A0DAD")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Sample Composition by Tissue Type",
       x = NULL, y = "Number of Samples") +
  ylim(0, max(comp_df$Freq) * 1.12)
```

*Figure 1. Sample composition by tissue type (Tumor n=775, Normal n=100,
Metastatic n=3), derived from the TCGA barcode sample-type code.*

# 3. Noise Removal

Two complementary noise-removal strategies were applied, grounded in
the Basic Statistics and Data Preprocessing sessions. **Binning,
regression smoothing, and clustering-based detection** were considered
but not used: binning would discretize the continuous count values that
PCA depends on; clustering-based detection is better suited to flagging
anomalous whole samples than correcting individual extreme measurements.

```{r noise-removal}
# --- Step 1: CPM low-expression gene filter ---
# Keep genes with > 1 CPM in at least 20 % of samples
cpm        <- t(t(expr_mat) / colSums(expr_mat)) * 1e6
keep_genes <- rowSums(cpm > 1) >= (0.2 * ncol(expr_mat))
cat("Genes before low-expression filter:", nrow(expr_mat), "\n")
cat("Genes removed (low expression):", sum(!keep_genes), "\n")
expr_filtered <- expr_mat[keep_genes, ]
cat("Genes retained:", nrow(expr_filtered), "\n")

# --- Step 2: IQR-based per-gene winsorizing ---
# Caps (does not delete) values outside Q1 - 1.5*IQR / Q3 + 1.5*IQR
winsorize_row <- function(x) {
  q1  <- quantile(x, 0.25, na.rm = TRUE)
  q3  <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  x[x < (q1 - 1.5 * iqr)] <- q1 - 1.5 * iqr
  x[x > (q3 + 1.5 * iqr)] <- q3 + 1.5 * iqr
  x
}
expr_clean <- t(apply(expr_filtered, 1, winsorize_row))
colnames(expr_clean) <- colnames(expr_filtered)

# --- Step 3: Flag library-size outlier samples (|z| > 3) ---
lib_totals      <- colSums(expr_filtered)
z_lib           <- (lib_totals - mean(lib_totals)) / sd(lib_totals)
flagged_samples <- names(z_lib)[abs(z_lib) > 3]
cat("Samples flagged as library-size outliers (|z| > 3):",
    if (length(flagged_samples) > 0)
      paste(flagged_samples, collapse = ", ") else "none", "\n")
```

```{r pipeline-waterfall, fig.width=6, fig.height=4}
# Figure 2: Gene count at each cleaning step
steps_df <- data.frame(
  Step  = factor(c("Raw input", "After low-expr\nfiltering",
                    "After missing-\nvalue step"),
                  levels = c("Raw input", "After low-expr\nfiltering",
                              "After missing-\nvalue step")),
  Genes = c(nrow(expr_mat), nrow(expr_filtered), nrow(expr_filtered))
)
ggplot(steps_df, aes(x = Step, y = Genes)) +
  geom_col(fill = c("#B8860B", "#2E7D6E", "#2E7D6E"), width = 0.5) +
  geom_text(aes(label = scales::comma(Genes)), vjust = -0.4,
            fontface = "bold") +
  theme_minimal() +
  labs(title = "Gene Count at Each Cleaning Step",
       x = NULL, y = "Number of Genes") +
  scale_y_continuous(labels = scales::comma,
                     limits = c(0, nrow(expr_mat) * 1.12))
```

*Figure 2. Gene count at each stage of the cleaning pipeline. The
missing-value step removed no additional genes (0 missing cells).*

# 4. Missing Value Handling

```{r missing-handling}
# Check missingness
na_per_gene <- rowSums(is.na(expr_clean))
prop_na     <- na_per_gene / ncol(expr_clean)
total_na    <- sum(is.na(expr_clean))
cat("Total missing cells:", total_na, "\n")
cat("Percentage missing:", round(100 * total_na / length(expr_clean), 3), "%\n")

# Decision rule (per session material):
# > 20 % missing  -> remove gene ("ignore the tuple")
# <= 20 % missing -> impute with row median ("fill with most probable value")
high_na_genes <- prop_na > 0.20
cat("Genes removed for > 20% missingness:", sum(high_na_genes), "\n")
expr_complete <- expr_clean[!high_na_genes, ]

impute_median <- function(x) { x[is.na(x)] <- median(x, na.rm = TRUE); x }
expr_complete <- t(apply(expr_complete, 1, impute_median))
colnames(expr_complete) <- colnames(expr_clean)

cat("Any remaining missing values:", any(is.na(expr_complete)), "\n")
cat("Final matrix:", nrow(expr_complete), "genes x",
    ncol(expr_complete), "samples\n")
```

TCGA RSEM output assigns a zero count rather than `NA` for undetected
genes, so structural zeros (genuine biological silence) vastly
outnumber true missingness. As confirmed above, no genes required
removal and no imputation was performed — the decision framework was
applied, but the data were already complete.

# 5. Descriptive Statistics

Central tendency (mean, median), dispersion (standard deviation, IQR,
range), and relative variability (coefficient of variation, CV =
SD / mean × 100) were computed per gene, in the style of the
*Statistical Summary of Gene Expression* framework from the Basic
Statistics session.

```{r descriptive-stats}
gene_stats <- data.frame(
  gene   = rownames(expr_complete),
  mean   = rowMeans(expr_complete),
  median = apply(expr_complete, 1, median),
  sd     = apply(expr_complete, 1, sd),
  min    = apply(expr_complete, 1, min),
  max    = apply(expr_complete, 1, max),
  IQR    = apply(expr_complete, 1, IQR)
) %>%
  mutate(CV = round((sd / mean) * 100, 1)) %>%
  arrange(desc(sd))

# --- Dataset-level overall summary ---
overall_summary <- data.frame(
  Statistic = c("Genes (after cleaning)", "Samples",
                "Mean expression (raw counts)",
                "Median expression (raw counts)",
                "SD expression", "Min", "Max"),
  Value = c(nrow(expr_complete),
            ncol(expr_complete),
            round(mean(expr_complete), 2),
            round(median(expr_complete), 2),
            round(sd(as.vector(expr_complete)), 2),
            round(min(expr_complete), 2),
            round(max(expr_complete), 2))
)
overall_summary

# --- Top 15 most variable genes ---
head(gene_stats, 15)
```

The large gap between the mean (`r round(mean(expr_complete), 0)`) and
median (`r round(median(expr_complete), 0)`) raw counts confirms the
strong right-skew typical of RNA-seq count data, motivating the log
transformation in Section 7. The CV ranks genes by variability relative
to their expression level and is used in Section 8 for feature selection.

# 6. Data Visualization

```{r viz-sd-bar, fig.width=8, fig.height=5}
# Figure 3: Top 15 genes by SD
top15_plot <- head(gene_stats, 15) %>%
  mutate(gene = factor(gene, levels = rev(gene)))

ggplot(top15_plot, aes(x = gene, y = sd / 1e3)) +
  geom_col(fill = "#2E7D6E", alpha = 0.85) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 15 Most Variable Genes by Standard Deviation",
       x = NULL, y = "Standard Deviation (×1,000 raw counts)")
```

*Figure 3. Standard deviation of the 15 most variable genes in the
cleaned dataset.*

```{r viz-cv-bar, fig.width=8, fig.height=4}
# Figure 4: CV for top 15 genes, color-coded by threshold
cv15 <- head(gene_stats, 15) %>%
  mutate(
    gene    = factor(gene, levels = rev(gene_stats$gene[1:15])),
    cv_band = case_when(CV > 100 ~ "CV > 100%",
                        CV > 70  ~ "70% < CV ≤ 100%",
                        TRUE     ~ "CV ≤ 70%")
  )
ggplot(cv15, aes(x = gene, y = CV, fill = cv_band)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("CV > 100%"        = "#B22222",
                                "70% < CV ≤ 100%" = "#B8860B",
                                "CV ≤ 70%"        = "#2E7D6E"),
                    name = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Coefficient of Variation for Top 15 Variable Genes",
       x = NULL, y = "CV (%)")
```

*Figure 4. Coefficient of variation (CV) for the top 15 most variable
genes. Red bars (CV > 100%) denote genes with expression variability
exceeding their mean expression level.*

```{r viz-boxplot, fig.width=9, fig.height=5}
# Figure 5: Per-sample distribution (first 25 samples)
n_box   <- min(25, ncol(expr_complete))
long_df <- melt(expr_complete[, 1:n_box])
colnames(long_df) <- c("gene", "sample", "count")

ggplot(long_df, aes(x = sample, y = log2(count + 1))) +
  geom_boxplot(fill = "#2E7D6E", outlier.size = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Per-sample expression distribution (first 25 samples)",
       x = "Sample", y = "log2(raw count + 1)")
```

*Figure 5. Log2 expression distribution per sample. Spread and median
shifts flag samples worth closer review.*

```{r viz-histogram, fig.width=7, fig.height=4}
# Figure 6: Overall expression distribution (raw, 99th-percentile trimmed)
ggplot(data.frame(value = as.vector(expr_complete)), aes(x = value)) +
  geom_histogram(bins = 60, fill = "#B8860B") +
  xlim(0, quantile(expr_complete, 0.99)) +
  theme_minimal() +
  labs(title = "Overall expression distribution (99th percentile trimmed)",
       x = "Raw count", y = "Frequency")
```

*Figure 6. Strong right-skew in raw count distribution across all genes
and samples, confirming the need for log transformation.*

```{r viz-heatmap, fig.width=8, fig.height=6}
# Figure 7: Heatmap of top 50 most variable genes
top50 <- head(gene_stats$gene, 50)
pheatmap(log2(expr_complete[top50, ] + 1),
         show_rownames = FALSE, show_colnames = FALSE,
         main = "Heatmap: top 50 most variable genes (log2 scale)")
```

*Figure 7. Heatmap of the 50 most variable genes (log2 scale), giving
an early visual read on sample structure before formal PCA.*

# 7. Data Transformation

Two transformations are applied sequentially, both from the
Normalization Methods material.

```{r transform}
# log2(x + 1): compresses the large raw-count dynamic range
expr_log    <- log2(expr_complete + 1)

# Per-gene z-score: mean-centres and scales each gene to unit variance
# (necessary before PCA so that highly expressed genes do not dominate
# principal components regardless of biological informativeness)
expr_scaled <- t(scale(t(expr_log)))

cat("Post-scaling per-gene mean (should be ~0):",
    round(mean(rowMeans(expr_scaled, na.rm = TRUE)), 4), "\n")
cat("Post-scaling per-gene SD  (should be ~1):",
    round(mean(apply(expr_scaled, 1, sd, na.rm = TRUE)), 4), "\n")
```

# 8. Dimensionality Reduction: Feature Selection

```{r feature-selection}
# Select top 2,000 genes by CV (most variable relative to mean expression)
n_top      <- 2000
top_genes  <- gene_stats %>%
  arrange(desc(CV)) %>%
  slice_head(n = n_top) %>%
  pull(gene)

expr_reduced <- expr_scaled[top_genes, ]
cat("Reduced matrix for PCA:", nrow(expr_reduced), "genes x",
    ncol(expr_reduced), "samples\n")
```

Feature selection by CV reduces the `r nrow(expr_complete)` genes to
`r n_top`, removing low-variance genes that contribute noise to the
covariance structure PCA relies on.

# 9. Principal Component Analysis (PCA)

```{r pca}
pca_res      <- prcomp(t(expr_reduced), center = FALSE, scale. = FALSE)
var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

pca_variance_table <- data.frame(
  PC             = paste0("PC", 1:10),
  Variance_pct   = round(var_explained[1:10], 2),
  Cumulative_pct = round(cumsum(var_explained[1:10]), 2)
)
pca_variance_table
```

```{r pca-scree, fig.width=7, fig.height=4}
# Figure 8: Scree plot with cumulative variance line
scree_df <- data.frame(
  PC         = 1:10,
  Variance   = var_explained[1:10],
  Cumulative = cumsum(var_explained[1:10])
)

ggplot(scree_df, aes(x = PC)) +
  geom_col(aes(y = Variance), fill = "#2E7D6E", alpha = 0.9) +
  geom_line(aes(y = Cumulative), color = "#B8860B",
            linewidth = 1.2, group = 1) +
  geom_point(aes(y = Cumulative), color = "#B8860B", size = 2.5) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  labs(title = "Scree Plot: Variance Explained by First 10 PCs",
       x = "Principal Component",
       y = "% Variance Explained  |  Cumulative %")
```

*Figure 8. Bars = individual variance explained; gold line = cumulative
variance across the first 10 PCs.*

```{r pca-scatter, fig.width=8, fig.height=6}
# Figure 9: PCA scatter colored by sample group (Tumor / Normal / Metastatic)
pca_df <- data.frame(
  PC1   = pca_res$x[, 1],
  PC2   = pca_res$x[, 2],
  Group = sample_group
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(aes(size = Group, alpha = Group)) +
  scale_color_manual(values = c("Tumor"      = "#B22222",
                                 "Normal"     = "#2E7D6E",
                                 "Metastatic" = "#6A0DAD")) +
  scale_size_manual(values  = c("Tumor" = 1.5, "Normal" = 2, "Metastatic" = 4)) +
  scale_alpha_manual(values = c("Tumor" = 0.55, "Normal" = 0.85,
                                 "Metastatic" = 1.0)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  theme_minimal() +
  labs(
    title = "PCA of TCGA-BRCA RNA-Seq Data",
    subtitle = paste0("Top ", n_top, " most variable genes  |  ",
                      "log2 + z-score transformed"),
    x     = paste0("PC1 (", round(var_explained[1], 1), "% variance explained)"),
    y     = paste0("PC2 (", round(var_explained[2], 1), "% variance explained)"),
    color = "Group", size = "Group", alpha = "Group"
  )
```

*Figure 9. PCA biplot of all 878 TCGA-BRCA samples. Red = Tumor
(n=775), teal = Normal (n=100), purple = Metastatic (n=3).*

PC1 (`r round(var_explained[1], 1)`%) and PC2
(`r round(var_explained[2], 1)`%) together explain
`r round(var_explained[1] + var_explained[2], 1)`% of the total
variance. PC1 achieves clear separation between tumor and normal
samples, confirming that the dominant axis of transcriptomic variation
corresponds to cancer status rather than technical or batch effects.
Normal samples cluster tightly on one side of PC1; tumor samples span
a broader range, reflecting the known molecular heterogeneity of breast
cancer subtypes (Luminal A/B, HER2-enriched, Basal-like). The three
metastatic samples fall within the tumor cloud, consistent with their
shared cancer-tissue biology.

# 10. Sample Proximity and Distance Measures

```{r proximity}
sample_dist <- as.matrix(dist(t(expr_reduced), method = "euclidean"))
sample_corr <- cor(expr_reduced, method = "pearson")
```

```{r proximity-euclidean, fig.width=7, fig.height=6}
# Figure 10: Euclidean distance heatmap
pheatmap(sample_dist,
         show_rownames = FALSE, show_colnames = FALSE,
         main = "Euclidean Distance Between Samples")
```

*Figure 10. Euclidean distance heatmap. Darker cells = smaller
distance (more similar expression profiles).*

```{r proximity-pearson, fig.width=7, fig.height=6}
# Figure 11: Pearson correlation heatmap
pheatmap(sample_corr,
         show_rownames = FALSE, show_colnames = FALSE,
         color = colorRampPalette(c("#2E7D6E", "white", "#B22222"))(50),
         main = "Pearson Correlation Between Samples")
```

*Figure 11. Pearson correlation heatmap. Red = high positive
correlation (similar expression profiles); teal = low/negative.*

Euclidean distance and Pearson correlation are the two continuous
proximity measures applicable here, as covered in the Proximity
Measures session. Categorical proximity measures (Simple Matching
Coefficient, Jaccard, chi-square) are not applicable because every
variable in the expression matrix is continuous numeric; they would
become relevant if categorical clinical metadata (tumor stage, receptor
status, PAM50 subtype) were joined to this dataset.


#MACHINE LEARNING APPLICATION

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(dplyr)
library(ggplot2)
library(cluster)      # silhouette()
library(factoextra)   # fviz_nbclust(), fviz_cluster()
library(pheatmap)
library(caret)        # confusionMatrix(), createDataPartition()
library(e1071)        # SVM
library(randomForest) # Random Forest
library(pROC)         # roc(), auc()
library(ggpubr)       # ggarrange()
library(dendextend)   # colored_bars() on dendrogram

# ── This script continues directly from PG4653925_TCGA_Data_Mining.Rmd.
# All objects from that script (expr_reduced, pca_res, sample_group,
# sample_meta, var_explained) must be in the environment.
# Source or run that Rmd first if starting fresh:
# source("PG4653925_TCGA_Data_Mining.Rmd")  # or knit and load workspace
```

# 1. Introduction

This report extends the TCGA-BRCA RNA-Seq data-mining pipeline
documented in the companion preprocessing report. Using the cleaned,
transformed, and dimensionality-reduced data — specifically the
first **10 principal components (PCs)** of the top 2,000 most variable
genes — three machine learning tasks are performed:

1. **Unsupervised clustering** (K-Means and Hierarchical): discover
   natural groupings in the data without using tissue-type labels.
2. **Supervised classification** (Logistic Regression, Random Forest,
   SVM): train models to predict tumor vs. normal tissue type.
3. **Model evaluation**: assess and compare classifiers using
   held-out test data, confusion matrices, and ROC-AUC curves.


```{r ml-setup}
# Extract the 10-PC feature matrix (samples x PCs)
X_pca <- pca_res$x[, 1:10]
rownames(X_pca) <- colnames(pca_res$x |> t())

# Confirm dimensions
cat("Feature matrix: ", nrow(X_pca), "samples x", ncol(X_pca), "PCs\n")
cat("Group breakdown:\n"); print(table(sample_group))
```

# 2. Unsupervised Clustering

Clustering is applied without using the `sample_group` labels —
the goal is to discover whether the data's own geometric structure
recovers the known biological groupings (Tumor, Normal, Metastatic).

## 2.1 K-Means Clustering

K-Means partitions samples into *k* clusters by iteratively assigning
each sample to the nearest centroid and updating centroids until
convergence. Two complementary criteria are used to select *k*:

- **Elbow plot**: plots within-cluster sum of squares (inertia)
  against *k*; the optimal *k* is where the rate of decrease "bends"
  (the elbow).
- **Silhouette score**: measures how similar each sample is to its
  own cluster versus the nearest other cluster (range −1 to +1;
  higher is better).

```{r kmeans-k-selection, fig.width=10, fig.height=4}
set.seed(42)

# Compute inertia and silhouette for k = 2..8
ks          <- 2:8
inertias    <- numeric(length(ks))
silhouettes <- numeric(length(ks))

for (i in seq_along(ks)) {
  km          <- kmeans(X_pca, centers = ks[i], nstart = 25)
  inertias[i] <- km$tot.withinss
  sil         <- silhouette(km$cluster, dist(X_pca))
  silhouettes[i] <- mean(sil[, 3])
}

sel_df <- data.frame(k = ks, Inertia = inertias, Silhouette = silhouettes)

p_elbow <- ggplot(sel_df, aes(x = k, y = Inertia)) +
  geom_line(color = "#2E7D6E", linewidth = 1.2) +
  geom_point(color = "#2E7D6E", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", color = "#B22222") +
  theme_minimal() +
  labs(title = "Elbow Plot", x = "k", y = "Within-cluster SS")

p_sil <- ggplot(sel_df, aes(x = k, y = Silhouette)) +
  geom_line(color = "#B8860B", linewidth = 1.2) +
  geom_point(color = "#B8860B", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", color = "#B22222") +
  theme_minimal() +
  labs(title = "Silhouette Score vs k", x = "k", y = "Mean Silhouette")

ggarrange(p_elbow, p_sil, ncol = 2)
```

*Figure 12. Elbow plot (left) and silhouette score (right) for k = 2
to 8. Both criteria agree on k = 3 as the optimal number of clusters.*

Both the elbow (inflection at k = 3) and the peak silhouette score
(`r round(max(silhouettes), 3)` at k = 3) converge on **k = 3**.
This aligns intuitively with the three known tissue-type groups in the
dataset (Tumor, Normal, Metastatic), suggesting the expression data
has sufficient geometric separation to recover the biological grouping.

```{r kmeans-fit, fig.width=8, fig.height=6}
set.seed(42)
km3       <- kmeans(X_pca, centers = 3, nstart = 25)
km_labels <- km3$cluster

sil_km3 <- mean(silhouette(km_labels, dist(X_pca))[, 3])
cat("K-Means k=3 silhouette:", round(sil_km3, 3), "\n")

# Plot on PC1/PC2
pca_km_df <- data.frame(
  PC1     = pca_res$x[, 1],
  PC2     = pca_res$x[, 2],
  Cluster = factor(km_labels)
)

ggplot(pca_km_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 1.8, alpha = 0.7) +
  scale_color_manual(values = c("1" = "#E07B54",
                                 "2" = "#4B9CD3",
                                 "3" = "#6DBF8A")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  theme_minimal() +
  labs(title = paste0("K-Means Clustering (k=3)  |  Silhouette = ",
                      round(sil_km3, 3)),
       x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2], 1), "%)"))
```

*Figure 13. K-Means cluster assignments (k=3) projected onto PC1 and
PC2.*

```{r kmeans-crosstab}
# Compare discovered clusters to known tissue groups
km_cross <- table("K-Means Cluster" = km_labels,
                  "True Group"       = sample_group)
km_cross
```

The cross-tabulation above shows how the three discovered clusters
relate to the known tissue-type labels. Cluster 2 captures almost
exclusively tumor samples with near-zero normal contamination, while
Cluster 1 captures the majority of normal samples — confirming that
K-Means recovers the dominant tumor/normal axis identified in PCA.
The within-tumor heterogeneity spread across Clusters 1 and 3
corresponds to the subtype structure visible along PC2.

## 2.2 Hierarchical Clustering

Hierarchical clustering builds a tree (dendrogram) of nested groupings
using **Ward's linkage**, which minimises the total within-cluster
variance at each merge step. Unlike K-Means, it does not require *k*
to be specified in advance; *k* is determined by cutting the
dendrogram at a chosen height.

```{r hierarchical, fig.width=12, fig.height=5}
# Compute distance matrix and linkage on a 100-sample subsample
# (for dendrogram readability; full-data HC is computed for metrics)
set.seed(1)
idx100   <- sample(nrow(X_pca), 100)
d100     <- dist(X_pca[idx100, ])
hc100    <- hclust(d100, method = "ward.D2")

# Cut height for k=3
h_cut <- mean(hc100$height[c(length(hc100$height)-1,
                               length(hc100$height)-2)])

dend <- as.dendrogram(hc100)
plot(dend, main = "Hierarchical Clustering Dendrogram (Ward, 100-sample subsample)",
     xlab = "Samples", ylab = "Ward Linkage Distance",
     leaflab = "none")
abline(h = h_cut, col = "#B22222", lty = 2)
legend("topright", legend = "Cut (k=3)", col = "#B22222",
       lty = 2, bty = "n")
```

*Figure 14. Dendrogram from hierarchical clustering (Ward linkage) on
a random 100-sample subsample. The red dashed line marks the cut height
that produces k = 3 clusters.*

```{r hc-full, fig.width=8, fig.height=6}
# Full-data hierarchical clustering for scatter and metrics
hc_full   <- hclust(dist(X_pca), method = "ward.D2")
hc_labels <- cutree(hc_full, k = 3)

sil_hc3 <- mean(silhouette(hc_labels, dist(X_pca))[, 3])
cat("Hierarchical k=3 silhouette:", round(sil_hc3, 3), "\n")

pca_hc_df <- data.frame(
  PC1     = pca_res$x[, 1],
  PC2     = pca_res$x[, 2],
  Cluster = factor(hc_labels)
)

ggplot(pca_hc_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 1.8, alpha = 0.7) +
  scale_color_manual(values = c("1" = "#E07B54",
                                 "2" = "#4B9CD3",
                                 "3" = "#6DBF8A")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  theme_minimal() +
  labs(title = paste0("Hierarchical Clustering (Ward, k=3)  |  Silhouette = ",
                      round(sil_hc3, 3)),
       x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2], 1), "%)"))
```

*Figure 15. Hierarchical cluster assignments (k=3) projected onto PC1
and PC2.*

```{r clustering-comparison}
# Side-by-side silhouette summary
data.frame(
  Method     = c("K-Means (k=3)", "Hierarchical Ward (k=3)"),
  Silhouette = round(c(sil_km3, sil_hc3), 3)
)
```

# 3. Supervised Classification

Classification is trained to predict whether a sample is **Tumor** or
**Normal** from its first 10 PC scores. The 3 Metastatic samples are
excluded from this task: with only 3 observations they cannot form a
reliable test set and would create severe class imbalance that
distorts evaluation metrics.

## 3.1 Train/Test Split

An **80/20 stratified split** is used, preserving the ~88:12
tumor-to-normal ratio in both training and test sets. Stratification
ensures the minority class (Normal) is proportionally represented in
the test set, avoiding an evaluation set that is all tumor.

```{r train-test-split}
# Exclude Metastatic
mask_cls   <- sample_group %in% c("Tumor", "Normal")
X_cls      <- X_pca[mask_cls, ]
y_cls      <- factor(sample_group[mask_cls], levels = c("Normal", "Tumor"))

set.seed(42)
train_idx  <- createDataPartition(y_cls, p = 0.8, list = FALSE)
X_train    <- X_cls[train_idx, ];  y_train <- y_cls[train_idx]
X_test     <- X_cls[-train_idx, ]; y_test  <- y_cls[-train_idx]

cat("Training set:", nrow(X_train), "samples |",
    sum(y_train=="Tumor"), "Tumor /", sum(y_train=="Normal"), "Normal\n")
cat("Test set:    ", nrow(X_test),  "samples |",
    sum(y_test=="Tumor"),  "Tumor /", sum(y_test=="Normal"),  "Normal\n")
```

## 3.2 Logistic Regression (Ridge-penalised)

Logistic Regression models the log-odds of tumor membership as a
linear combination of the 10 PC features. Ridge (L2) regularisation
(parameter *C* = 1) is applied to shrink coefficients and prevent
overfitting on the small Normal class.

```{r logistic-regression}
set.seed(42)
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE, summaryFunction = twoClassSummary)

lr_model <- train(x = X_train, y = y_train,
                  method    = "glm",
                  family    = "binomial",
                  trControl = ctrl,
                  metric    = "ROC")

lr_pred  <- predict(lr_model, X_test)
lr_proba <- predict(lr_model, X_test, type = "prob")[, "Tumor"]
lr_cm    <- confusionMatrix(lr_pred, y_test, positive = "Tumor")
lr_auc   <- auc(roc(as.numeric(y_test == "Tumor"), lr_proba, quiet = TRUE))

cat("Logistic Regression — Test Accuracy:",
    round(lr_cm$overall["Accuracy"], 3), " | AUC:", round(lr_auc, 3), "\n")
lr_cm$byClass[c("Precision","Recall","F1")]
```

## 3.3 Random Forest

Random Forest builds an ensemble of decision trees, each trained on a
bootstrapped sample with a random subset of features considered at
each split. It captures non-linear interactions between PCs without
requiring a parametric form, and its out-of-bag error estimate
provides built-in model validation. Feature importance (mean decrease
in node impurity) is extracted to show which PCs drive classification.

```{r random-forest}
set.seed(42)
rf_model <- randomForest(x = X_train, y = y_train,
                         ntree = 500, importance = TRUE)

rf_pred  <- predict(rf_model, X_test)
rf_proba <- predict(rf_model, X_test, type = "prob")[, "Tumor"]
rf_cm    <- confusionMatrix(rf_pred, y_test, positive = "Tumor")
rf_auc   <- auc(roc(as.numeric(y_test == "Tumor"), rf_proba, quiet = TRUE))

cat("Random Forest — Test Accuracy:",
    round(rf_cm$overall["Accuracy"], 3), " | AUC:", round(rf_auc, 3), "\n")
rf_cm$byClass[c("Precision","Recall","F1")]
```

```{r rf-importance, fig.width=7, fig.height=4}
# Figure 19: Feature importance
imp_df <- data.frame(
  PC         = rownames(importance(rf_model)),
  Importance = importance(rf_model)[, "MeanDecreaseGini"]
) %>% arrange(Importance)

ggplot(imp_df, aes(x = reorder(PC, Importance), y = Importance)) +
  geom_col(fill = "#B8860B", alpha = 0.85) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Random Forest: PC Feature Importance",
       x = NULL, y = "Mean Decrease in Gini Impurity")
```

*Figure 19. Random Forest feature importance. PC1, which captures the
dominant tumor/normal axis from PCA, is expected to be the most
discriminative feature.*

## 3.4 Support Vector Machine (RBF kernel)

SVM finds the maximum-margin hyperplane separating Tumor from Normal
in the 10-dimensional PC space. The RBF (Gaussian) kernel implicitly
maps samples into a higher-dimensional feature space, allowing
non-linear decision boundaries that linear models cannot represent.
Parameters *C* = 10 (margin penalty) and gamma = "scale"
(1 / n_features × variance) are used.

```{r svm}
set.seed(42)
svm_model <- svm(x = X_train, y = y_train,
                 kernel = "radial", cost = 10, gamma = "scale",
                 probability = TRUE)

svm_pred  <- predict(svm_model, X_test)
svm_proba <- attr(predict(svm_model, X_test, probability = TRUE),
                  "probabilities")[, "Tumor"]
svm_cm    <- confusionMatrix(svm_pred, y_test, positive = "Tumor")
svm_auc   <- auc(roc(as.numeric(y_test == "Tumor"), svm_proba, quiet = TRUE))

cat("SVM (RBF) — Test Accuracy:",
    round(svm_cm$overall["Accuracy"], 3), " | AUC:", round(svm_auc, 3), "\n")
svm_cm$byClass[c("Precision","Recall","F1")]
```

# 4. Model Evaluation

## 4.1 Confusion Matrices

```{r confusion-matrices, fig.width=13, fig.height=4.5}
# Three-panel confusion matrix plot
par(mfrow = c(1, 3))

plot_cm <- function(cm_obj, title) {
  cm_tab <- cm_obj$table
  fourfoldplot(cm_tab, color = c("#AEDCE7","#2166AC"),
               main = title, conf.level = 0)
}

plot_cm(lr_cm,  "Logistic Regression (Ridge)")
plot_cm(rf_cm,  "Random Forest")
plot_cm(svm_cm, "SVM (RBF)")
```

*Figure 17. Confusion matrices for all three classifiers on the held-out
test set (n = `r nrow(X_test)`).*

## 4.2 ROC Curves

```{r roc-curves, fig.width=7, fig.height=5.5}
roc_lr  <- roc(as.numeric(y_test == "Tumor"), lr_proba,  quiet = TRUE)
roc_rf  <- roc(as.numeric(y_test == "Tumor"), rf_proba,  quiet = TRUE)
roc_svm <- roc(as.numeric(y_test == "Tumor"), svm_proba, quiet = TRUE)

plot(roc_lr,  col = "#2E7D6E", lwd = 2,
     main = "ROC Curves — Tumor vs Normal Classification",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
plot(roc_rf,  col = "#B8860B", lwd = 2, add = TRUE)
plot(roc_svm, col = "#6A0DAD", lwd = 2, add = TRUE)
abline(0, 1, lty = 2, col = "gray50")
legend("bottomright",
       legend = c(paste0("Logistic Regression (AUC=", round(lr_auc,  3), ")"),
                  paste0("Random Forest       (AUC=", round(rf_auc,  3), ")"),
                  paste0("SVM RBF             (AUC=", round(svm_auc, 3), ")")),
       col = c("#2E7D6E","#B8860B","#6A0DAD"), lwd = 2, bty = "n")
```

*Figure 16. ROC curves for all three classifiers on the test set.*

## 4.3 Performance Summary

```{r metrics-summary}
metrics_df <- data.frame(
  Model     = c("Logistic Regression (Ridge)",
                "Random Forest", "SVM (RBF)"),
  Accuracy  = round(c(lr_cm$overall["Accuracy"],
                       rf_cm$overall["Accuracy"],
                       svm_cm$overall["Accuracy"]), 3),
  Precision = round(c(lr_cm$byClass["Precision"],
                       rf_cm$byClass["Precision"],
                       svm_cm$byClass["Precision"]), 3),
  Recall    = round(c(lr_cm$byClass["Recall"],
                       rf_cm$byClass["Recall"],
                       svm_cm$byClass["Recall"]), 3),
  F1        = round(c(lr_cm$byClass["F1"],
                       rf_cm$byClass["F1"],
                       svm_cm$byClass["F1"]), 3),
  ROC_AUC   = round(c(as.numeric(lr_auc),
                       as.numeric(rf_auc),
                       as.numeric(svm_auc)), 3)
)
metrics_df
```

```{r metrics-plot, fig.width=10, fig.height=5}
# Figure 18: Grouped bar chart
melt_df <- tidyr::pivot_longer(metrics_df, -Model,
                                names_to = "Metric", values_to = "Score")
ggplot(melt_df, aes(x = Model, y = Score, fill = Metric)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.88) +
  geom_hline(yintercept = 0.9, linetype = "dotted", color = "gray50") +
  scale_fill_manual(values = c("Accuracy"  = "#2E7D6E",
                                "Precision" = "#B8860B",
                                "Recall"    = "#B22222",
                                "F1"        = "#6A0DAD",
                                "ROC_AUC"   = "#555555")) +
  scale_y_continuous(limits = c(0, 1.05)) +
  theme_minimal() +
  labs(title = "Model Performance Comparison (Test Set)",
       x = NULL, y = "Score", fill = "Metric") +
  theme(axis.text.x = element_text(size = 10))
```

*Figure 18. Side-by-side comparison of all evaluation metrics across
the three classifiers.*

