---
title: "16S Pipeline — Package Setup, QC, Filter & Trim"
subtitle: "16S rRNA Microbiome Pipeline (DADA2)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

Download the 16S reference genome (to be used for taxonomy here: [**https://zenodo.org/records/4587955**](https://zenodo.org/records/4587955){.uri} (SILVA v138.1)) *silva_nr99_v138.1_train_set.fa.gz* - train data *silva_species_assignment_v138.1.fa.gz* - species data

## Package Installation

```{r install-packages}
# CRAN packages
install.packages(c(
  "ggplot2", "devtools", "vegan", "dplyr",
  "tidyverse", "remotes", "ggpubr", "rstatix",
  "stringr", "data.table", "lubridate"
))

if (!requireNamespace("BiocManager", quietly = TRUE))
{
  install.packages("BiocManager")
}
BiocManager::install(c("dada2", "phyloseq", "DESeq2",
                       "ANCOMBC", "Biostrings", "microbiome"))

# nicer plots
devtools::install_github("thomasp85/patchwork")
remotes::install_github("gavinsimpson/ggvegan")
```

## Config paths

```{r config}
library(dada2)
library(data.table)
packageVersion("dada2")

# Paths
# Raw, gzipped FASTQs
data <- "~/Desktop/mini_desktop/KNUST/second_sem/high_throughput_seq/mouse_full_data/"

# project folder for all derived outputs
proj_dir <- "~//Desktop/mini_desktop/KNUST/second_sem/high_throughput_seq/mouse_full_data/"

# r object file dir
rds_dir <- file.path(proj_dir, "rds")

# filtered fastqs written by DADA2
filt_dir <- file.path(data, "filtered")

# SILVA training data files(for taxonomy)
ref_dir <- file.path(proj_dir, "reference")

for (d in c(proj_dir, rds_dir, ref_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

# SILVA Reference DB
silva_train <- file.path(ref_dir, "silva_nr99_v138.1_train_set.fa.gz")
silva_species <- file.path(ref_dir, "silva_species_assignment_v138.1.fa.gz")

# Threading (for memory and faster processing)
N_THREADS <- max(1, parallel::detectCores() - 1) # for filtering/denoising
TAX_THREADS <- 2 # keep low for taxonomy genx
TAX_CHUNK <- 2000 # ASVs per taxonomy batch

# Helper: report current R memory use
mem_report <- function(tag = "") {
  gc(full = TRUE)
  used_mb <- sum(gc()[, "used"] * c(8,8)) / 1e6
  cat(sprintf("[MEM %s] R heap ~%.0f MB\n", tag, used_mb))
  invisible(NULL)
}
```

## The Dataset

This is the full **Schloss lab** mouse-gut dataset with the following convention

```         
M1D13_S12_L001_R1_001.fastq.gz
└───┘ └─┘ └──┘ └┘ └─┘
  │    │    │   │   └ read chunk index
  │    │    │   └──── read direction (R1 = forward, R2 = reverse)
  │    │    └──────── lane
  │    └───────────── Illumina sample number (S12)
  └────────────────── sample id (M1D13)
```

## Read in and sort the fastq file names

```{r list-fastqs}
# forward (R1) and reverse (R2) reads
dataF <- sort(list.files(data, pattern = "_R1_001\\.fastq\\.gz$", full.names = TRUE))
dataR <- sort(list.files(data, pattern = "_R2_001\\.fastq\\.gz$", full.names = TRUE))

# sample ID
sample.names <- sapply(strsplit(basename(dataF), "_"), `[`, 1)
cat("Forward files:", length(dataF)," Reverse files:", length(dataR), "\n")
cat("First few samples:", paste(head(sample.names), collapse = ", "), "\n")

# sanity checks
stopifnot(length(dataF) == length(dataR))
# confirm R1/R2 line up sample-for-sample
ok <- identical(
  sub("_R1_001\\.fastq\\.gz$", "", basename(dataF)),
  sub("_R2_001\\.fastq\\.gz$", "", basename(dataR))
)
stopifnot(ok)
```

## Step 1: Inspect Read Quality Profiles

```{r qc-forward}
# aggregate profile across a random subset of samples
set.seed(100)
idx <- sort(sample(seq_along(dataF), min(20, length(dataF))))

plotQualityProfile(dataF[idx], aggregate = TRUE) + ggplot2::ggtitle("Forward (R1) - aggregate")
```

```{r qc-reverse}
plotQualityProfile(dataR[idx], aggregate = TRUE) + ggplot2::ggtitle("Reverse (R2) - aggregate")
```

## Step 2: Filter and Trim

```{r filter-paths}
filt.dataF <- file.path(filt_dir, paste0(sample.names, "_F_filt.fastq.gz"))
filt.dataR <- file.path(filt_dir, paste0(sample.names, "_R_filt.fastq.gz"))

names(filt.dataF) <- sample.names
names(filt.dataR) <- sample.names
```

| Parameter | Forward | Reverse | Meaning |
|-------------------|------------------|------------------|------------------|
| `truncLen` | 240 | 160 | Truncate reads at this length (guided by quality plots) |
| `maxEE` | 2 | 2 | Max expected errors per read |
| `maxN` | 0 | 0 | No ambiguous bases allowed |
| `truncQ` | 2 | 2 | Truncate at first base with quality ≤ 2 |
| `rm.phix` | TRUE | TRUE | Remove PhiX control reads |

```{r filter-and-trim}
out <- filterAndTrim(
  dataF, filt.dataF, dataR, filt.dataR,
  truncLen = c(240, 160),
  maxN = 0, maxEE = c(2, 2), truncQ = 2, rm.phix = TRUE,
  compress = TRUE, multithread = N_THREADS
)
head(out)

# some samples can lose ALL reads after filtering which can break downstream steps - drop them
kept <- file.exists(filt.dataF) &
  file.exists(filt.dataR)
if (any(!kept)) {
  message("Dropping ", sum(!kept), " sample(s) with no reads after filtering: ",
          paste(sample.names[!kept], collapse = ", "))
}
filt.dataF <- filt.dataF[kept]
filt.dataR <- filt.dataR[kept]
sample.names <- sample.names[kept]

saveRDS(out, file.path(rds_dir, "filter_out.rds"))
saveRDS(sample.names, file.path(rds_dir, "sample_names.rds"))
```

```{r qc-after-filter}
plotQualityProfile(filt.dataR[idx], aggregate = TRUE)
plotQualityProfile(filt.dataF[idx], aggregate = TRUE)
```
