---
title: "16S Pipeline — Error Learning, Denoise/Merge, Chimera Removal, Taxonomy"
subtitle: "16S rRNA Microbiome Pipeline (DADA2) — group project"
output: html_document
---

## Step 3: Learn the Error Rates

```{r learn-errors}
errF <- learnErrors(filt.dataF, multithread = N_THREADS)
errR <- learnErrors(filt.dataR, multithread = N_THREADS)

saveRDS(errF, file.path(rds_dir, "errF.rds"))
saveRDS(errR, file.path(rds_dir, "errR.rds"))
mem_report("after learnErrors")
```

```{r plot-errors}
plotErrors(errF, nominalQ = TRUE)
plotErrors(errR, nominalQ = TRUE)
```

## Steps 4 - 5: Denoise + Merge

This is a memory-critical step and requires optimization. Instead of calling `dada2()` on **all** samples at once, loop **one sample at a time**: `dereplicate -> denoise F -> denoise R -> merge -> store only merged result -> free everything else -> gc()`

```{r denoise-merge-loop}
# (Re)load inputs - this block can be rerun in case session ends due to memory depletion 
sample.names <- readRDS(file.path(rds_dir, "sample_names.rds")) 
errF <- readRDS(file.path(rds_dir, "errF.rds")) 
errR <- readRDS(file.path(rds_dir, "errR.rds"))
filt.dataF <- file.path(filt_dir, paste0(sample.names, "_F_filt.fastq.gz")); names(filt.dataF) <- sample.names
filt.dataR <- file.path(filt_dir, paste0(sample.names, "_R_filt.fastq.gz")); names(filt.dataR) <- sample.names

mergers <- vector("list", length(sample.names)); names(mergers) <- sample.names

denoisedF <- integer(length(sample.names)); names(denoisedF) <- sample.names
denoisedR <- integer(length(sample.names)); names(denoisedR) <- sample.names

getN <- function(x) sum(getUniques(x))

for (sam in sample.names) {
  cat("Denoising:", sam, "\n")
  
  derepF <- derepFastq(filt.dataF[[sam]])
  ddF <- dada(derepF, err = errF, multithread = N_THREADS)
  
  derepR <- derepFastq(filt.dataR[[sam]])
  ddR <- dada(derepR, err = errR, multithread = N_THREADS)
  
  merger <- mergePairs(ddF, derepF, ddR, derepR, verbose = FALSE)
  
  mergers[[sam]] <- merger
  denoisedF[[sam]] <- getN(ddF)
  denoisedR[[sam]] <- getN(ddR)
  
  # free per-sample objects before the next iteration
  rm(derepF, derepR, ddF, ddR, merger)
  gc(full = TRUE)
}

saveRDS(mergers, file.path(rds_dir, "mergers.rds"))
saveRDS(denoisedF, file.path(rds_dir, "denoisedF.rds"))
saveRDS(denoisedR, file.path(rds_dir, "denoisedR.rds"))

mem_report("after denoise+merge loop")
```

## Step 6: Make the Sequence Table

```{r make-seqtab}
mergers <- readRDS(file.path(rds_dir, "mergers.rds"))

seqtab <- makeSequenceTable(mergers)
cat("Samples x ASVs:", paste(dim(seqtab), collapse = " x "), "\n")
print(table(nchar(getSequences(seqtab))))

saveRDS(seqtab, file.path(rds_dir, "seqtab.rds"))
```

## Step 7: Remove Chimeras

```{r remove-chimeras}
seqtab <- readRDS(file.path(rds_dir, "seqtab.rds"))

seqtab.nochim <- removeBimeraDenovo(
  seqtab, method = "consensus", multithread = N_THREADS, verbose = TRUE
)

cat("Non-chimeric ASVs:", ncol(seqtab.nochim), "\n")
cat("Fraction of reads kept:", round(sum(seqtab.nochim) / sum(seqtab), 4), "\n")

saveRDS(seqtab.nochim, file.path(rds_dir, "seqtab_nochim.rds"))
```

## Step 8: Track Reads the Pipeline

```{r track-reads}
out <- readRDS(file.path(rds_dir, "filter_out.rds"))
sample.names <- readRDS(file.path(rds_dir, "sample_names.rds"))
denoisedF <- readRDS(file.path(rds_dir, "denoisedF.rds"))
denoisedR <- readRDS(file.path(rds_dir, "denoisedR.rds"))
mergers <- readRDS(file.path(rds_dir, "mergers.rds"))
seqtab.nochim <- readRDS(file.path(rds_dir, "seqtab_nochim.rds"))

getM <- function(x) sum(getUniques(x))

# Align the filter matrix to surviving samples
out2 <- out[paste0(sample.names, "_F_filt.fastq.gz") %in% basename(names(denoisedF)) | TRUE, , drop = FALSE]

# rebuild from saved pieces matched by sample_id
track <- cbind(
  input = out[, "reads.in"][seq_along(sample.names)],
  filtered = out[, "reads.out"][seq_along(sample.names)],
  denoisedF = denoisedF,
  denoisedR = denoisedR,
  merged = sapply(mergers, getN),
  nonchim = rowSums(seqtab.nochim)[sample.names]
)
rownames(track) <- sample.names
head(track, 10)

saveRDS(track, file.path(rds_dir, "track_reads.rds"))
```

## Step 9: Assign Taxonomy - **chunked + low-thread**

```{r assign-taxonomy-chunked}
seqtab.nochim <- readRDS(file.path(rds_dir, "seqtab_nochim.rds"))
stopifnot(file.exists(silva_train)) # check that 16S ref is present

seqs <- getSequences(seqtab.nochim)
chunks <- split(seqs, ceiling(seq_along(seqs) / TAX_CHUNK))
cat("Assigning taxonomy for", length(seqs), "ASVs in", length(chunks), "chunk(s)\n")

taxa_chunks <- vector("list", length(chunks))
for (i in seq_along(chunks)) {
  cat(sprintf(" chunk %d / %d (%d ASVs)\n", i, length(chunks), length(chunks[[i]])))
  taxa_chunks[[i]] <- assignTaxonomy(
    chunks[[i]], silva_train,
    multithread = TAX_THREADS,
    tryRC = TRUE # guards against reverse-complement orientation issues
  )
  gc(full = TRUE) # release the reference + chunk before the next batch
}

taxa <- do.call(rbind, taxa_chunks)
rm(taxa_chunks); gc(full = TRUE)

# Species assignment - only run this if you have the species files
if (file.exists(silva_species)) {
  sp_chunks <- split(rownames(taxa), ceiling(seq_len(nrow(taxa)) / TAX_CHUNK))
  taxa_sp <- vector("list", length(sp_chunks))
  for (i in seq_along(sp_chunks)) {
    sub <- taxa[sp_chunks[[i]], , drop = FALSE]
    taxa_sp[[i]] <- addSpecies(sub, silva_species)
    gc(full = TRUE)
  }
  taxa <- do.call(rbind, taxa_sp)
  rm(taxa_sp); gc(full = TRUE)
}

saveRDS(taxa, file.path(rds_dir, "taxa.rds"))
mem_report("after taxonomy")
```

```{r taxa-inspect}
taxa.print <- taxa
rownames(taxa.print) <- NULL
head(taxa.print)

table(taxa[, "Phylum"], useNA = "ifany")
cat("ASVs with a species-level call:", sum(!is.na(taxa[, "Species"])), "of", nrow(taxa), "\n")
```
