---
title: "16S Pipeline — Differential Abundance (DESeq2 & ANCOM-BC2)"
subtitle: "16S rRNA Microbiome Pipeline (DADA2) — group project"
output: html_document
---

## Step 11: Differential Abundance Analysis

**Goal:** find ASVs/taxa whose abundance differs between *Early* and *Late* samples. Two complementary methods will be used: - **DESeq2**: Fast, familiar, integrates directly with the `ps` object

- **ANCOM-BC2**: compositionally-aware method with bias correction. Generally more appropriate default for microbiome relative-abundance data *NB: agreement between methods is reassuring*

```{r da-subset}
# build a clean data subset - drop controls and any empty ASVs
ps <- readRDS(file.path(rds_dir, "phyloseq.rds"))

ps_da <- subset_samples(ps, !is.na(When))
ps_da <- prune_taxa(taxa_sums(ps_da) > 0, ps_da)

# prevalence filter: keep ASVs seen in >= 10% of samples - remove ultra-rare taxa that are just noise
prev <- apply(otu_table(ps_da), 2, function(x) sum(x > 0))
keep <- prev >= max(2, ceiling(0.10 * nsamples(ps_da)))
ps_da <- prune_taxa(taxa_names(ps_da)[keep], ps_da)

# fix reference level so positive results = enriched in "Late"
sample_data(ps_da)$When <- factor(sample_data(ps_da)$When, levels = c("Early", "Late"))

ps_da
```

## 11a. DESeq2

```{r da-deseq2}
library(DESeq2)

# Gaurd against empty samples / taxa left behind by the prevalence filter
ps_da <- prune_samples(sample_sums(ps_da) > 0, ps_da)
ps_da <- prune_taxa(taxa_sums(ps_da) > 0, ps_da)

dds <- phyloseq_to_deseq2(ps_da, ~ When)

dds <- estimateSizeFactors(dds, type = "poscounts")

# sanity check - no NA size factors now
stopifnot(all(!is.na(sizeFactors(dds))))

dds <- DESeq(dds, test = "Wald", fitType = "local")
res <- results(dds, cooksCutoff = FALSE)

alpha <- 0.05
sigtab <- as.data.frame(res)
sigtab <- cbind(sigtab, as(tax_table(ps_da)[rownames(sigtab), ], "matrix"))
sigtab <- sigtab[!is.na(sigtab$padj) & sigtab$padj < alpha, ]
sigtab <- sigtab[order(sigtab$log2FoldChange), ]

cat("DESeq2: significant ASVs at padj <", alpha, ":", nrow(sigtab), "\n")
head(sigtab[, c("baseMean", "log2FoldChange", "padj", "Phylum", "Family", "Genus")], 20)

saveRDS(res, file.path(rds_dir, "deseq2_results.rds"))
saveRDS(sigtab, file.path(rds_dir, "deseq2_sigtab.rds"))
```

```{r da-deseq2-plot, fig.width=9, fig.height=6}
if (nrow(sigtab) > 0) {
  sigtab$Genus <- factor(sigtab$Genus, levels = unique(sigtab$Genus[order(sigtab$log2FoldChange)]))
  ggplot(sigtab, aes(x = log2FoldChange, y = Genus, color = Phylum)) + geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") + geom_point(size = 3, alpha = 0.8) + labs(title = "DESeq2: Early vs Late (positive = enriched in Late)", x = "log2 fold change", y = "Genus") + theme(axis.text.y = element_text(size = 8))
}
```

## 11b. ANCOM-BC2

ANCOM-BC2 models the sampling fraction bias inherent to compositional data, which DESeq2 does not. Run at the **Genus** level here (aggregating ASVs to genus is common for DA and reduces multiple-testing burden); To test ASVs directly, set "tax_level = NULL"

```{r da-ancombc2}
library(ANCOMBC)
library(microbiome)

set.seed(123)
anc <- ancombc2(
  data = ps_da,
  tax_level = "Genus",
  fix_formula = "When",
  group = "When",
  p_adj_method = "BH",
  prv_cut = 0.10, # prevalence filter inside ANCOM-BC
  struc_zero = TRUE,
  n_cl = TAX_THREADS,
  verbose = TRUE
)

anc_res <- anc$res
# effect/q-value cols are named after the contrast, e.g. "WhenLate"
lfc_col <- grep("^lfc_When", names(anc_res), value = TRUE)
q_col <- grep("^q_When", names(anc_res), value = TRUE)
diff_col <- grep("^diff_When", names(anc_res), value = TRUE)
anc_sig <- anc_res[anc_res[[diff_col]] == TRUE, c("taxon", lfc_col, q_col)]
anc_sig <- anc_sig[order(anc_sig[[lfc_col]]), ]

cat("ANCOM-BC2: differentially abundant genera:", nrow(anc_sig), "\n")
head(anc_sig, 20)

saveRDS(anc, file.path(rds_dir, "ancombc2_full.rds"))
saveRDS(anc_sig, file.path(rds_dir, "ancombc2_sig.rds"))
```

```{r da-ancombc2-lot, fig.width=9, fig.height=6}
if (nrow(anc_sig) > 0) {
  anc_sig$taxon <- factor(anc_sig$taxon, levels = anc_sig$taxon[order(anc_sig[[lfc_col]])])
  ggplot(anc_sig, aes(x = .data[[lfc_col]], y = taxon)) + geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") + geom_col(fill = "steelblue", alpha = 0.85) +
    labs(title = "ANCOM-BC2: Early vs Late (positive = enriched in Late)", x = "Bias-corrected log gold change", y = "Genus")
}
```

## 11c. Cross-check the two methods

> **Interpretation.** Taxa flagged by **both** methods are your most robust hits — report those with the most confidence. Method-specific hits are worth mentioning but should be flagged as less certain. ANCOM-BC2 is generally the more defensible primary result for compositional data; DESeq2 serves as a familiar sanity check.

```{r da-compare}
# genera flagged by DESeq2 (mapped from ASV-level results)
deseq_genera <- unique(na.omit(sigtab$Genus))
ancom_genera <- if (nrow(anc_sig))
  as.character(anc_sig$taxon) else character(0)

cat("DESEq2 genera:", length(deseq_genera), 
    "| ANCOM-BC2 genera:", length(ancom_genera),
    "| overlap:", length(intersect(deseq_genera, ancom_genera)), "\n")
intersect(deseq_genera, ancom_genera)
```

# Session Info
```{r session-info}
sessionInfo()
```
