---
title: "TCGA-BRCA RNA-Seq Data Mining: Cleaning, Exploration, and Dimensionality Reduction"
author: "Ebenezer Oppong Gyamfi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(readr)

#LOADING TCGA DATA
tcga_data <- read_tsv("biodata/BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt")

head(tcga_data)
```

# 1. Introduction

This report documents the complete data-mining workflow applied to the
TCGA Breast Invasive Carcinoma (BRCA) RNA-Seq Level 3 mRNAseq dataset
(RSEM output, Broad GDAC Firehose pipeline). The dataset encompasses
**878 samples** and an initial **20,503 gene identifiers**. The pipeline
covers: data structuring, noise removal, missing-value handling,
descriptive statistics, visualization, transformation, dimensionality
reduction, Principal Component Analysis (PCA), and sample proximity
measures — following the framework from the Data Preprocessing, Basic
Statistics, and Proximity Measures sessions. Every method is chosen
deliberately and explained in place.

# 2. Data Structuring

The raw TCGA export stores three value columns per sample (raw counts,
median-length-normalized counts, RPKM) under repeated sample barcodes,
and encodes each gene as `SYMBOL|ENTREZ_ID`. R appends `...N` suffixes
to deduplicate the repeated column headers, and the true value-type
labels live in row 1 of the data rather than in the header row. Three
structural operations were therefore required before analysis.

```{r structure-data}
# Step 1: Recover true barcodes by stripping R's auto-added "...N" suffix
barcodes     <- sub("\\.\\.\\.\\d+$", "", colnames(tcga_data))
barcodes[1]  <- "gene"
value_types  <- as.character(tcga_data[1, ])
value_types[1] <- ""
new_names    <- ifelse(value_types == "",
                       barcodes,
                       paste(barcodes, value_types, sep = "_"))
colnames(tcga_data) <- new_names

# Drop the embedded header row
expr_raw <- tcga_data[-1, ]

# Step 2: Split "SYMBOL|ENTREZ" gene identifier
expr_raw <- expr_raw %>%
  separate(gene, into = c("symbol", "entrez_id"), sep = "\\|", remove = FALSE)

# Step 3: Remove unmapped genes (symbol == "?")
n_before <- nrow(expr_raw)
expr_raw  <- expr_raw %>% filter(symbol != "?")
cat("Unmapped ('?') genes removed:", n_before - nrow(expr_raw),
    "out of", n_before, "\n")
```

```{r select-counts}
# Work with raw_counts only
counts_df  <- expr_raw %>% select(gene, symbol, entrez_id,
                                   ends_with("_raw_counts"))
count_cols <- grep("_raw_counts$", colnames(counts_df), value = TRUE)
counts_df[count_cols] <- lapply(counts_df[count_cols], as.numeric)

# Step 4: Collapse duplicate gene symbols — keep most-expressed row per symbol
n_dup <- sum(duplicated(counts_df$symbol) |
             duplicated(counts_df$symbol, fromLast = TRUE))
cat("Rows involved in duplicate gene symbols:", n_dup, "\n")

counts_df <- counts_df %>%
  mutate(row_total = rowSums(across(all_of(count_cols)), na.rm = TRUE)) %>%
  group_by(symbol) %>%
  slice_max(row_total, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-row_total)

# Build the numeric expression matrix
expr_mat <- as.matrix(counts_df[, count_cols])
rownames(expr_mat) <- counts_df$symbol
colnames(expr_mat) <- sub("_raw_counts$", "", count_cols)

cat("Matrix after structuring:", nrow(expr_mat), "genes x",
    ncol(expr_mat), "samples\n")
```

```{r sample-metadata}
# Derive sample group from TCGA barcode (positions 14-15 = sample-type code)
# 01-09 = tumor, 06 = metastatic, 10-19 = normal
sample_type_code <- substr(colnames(expr_mat), 14, 15)
sample_group <- case_when(
  as.numeric(sample_type_code) == 6              ~ "Metastatic",
  as.numeric(sample_type_code) < 10             ~ "Tumor",
  TRUE                                           ~ "Normal"
)
sample_meta <- data.frame(
  sample = colnames(expr_mat),
  type_code = sample_type_code,
  group = sample_group
)
table(sample_meta$group)
```

```{r sample-composition-plot, fig.width=5, fig.height=4}
# Figure 1: Sample composition
comp_df <- as.data.frame(table(Group = sample_meta$group))
ggplot(comp_df, aes(x = Group, y = Freq, fill = Group)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = Freq), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c("Tumor"      = "#B22222",
                                "Normal"     = "#2E7D6E",
                                "Metastatic" = "#6A0DAD")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Sample Composition by Tissue Type",
       x = NULL, y = "Number of Samples") +
  ylim(0, max(comp_df$Freq) * 1.12)
```

*Figure 1. Sample composition by tissue type (Tumor n=775, Normal n=100,
Metastatic n=3), derived from the TCGA barcode sample-type code.*

# 3. Noise Removal

Two complementary noise-removal strategies were applied, grounded in
the Basic Statistics and Data Preprocessing sessions. **Binning,
regression smoothing, and clustering-based detection** were considered
but not used: binning would discretize the continuous count values that
PCA depends on; clustering-based detection is better suited to flagging
anomalous whole samples than correcting individual extreme measurements.

```{r noise-removal}
# --- Step 1: CPM low-expression gene filter ---
# Keep genes with > 1 CPM in at least 20 % of samples
cpm        <- t(t(expr_mat) / colSums(expr_mat)) * 1e6
keep_genes <- rowSums(cpm > 1) >= (0.2 * ncol(expr_mat))
cat("Genes before low-expression filter:", nrow(expr_mat), "\n")
cat("Genes removed (low expression):", sum(!keep_genes), "\n")
expr_filtered <- expr_mat[keep_genes, ]
cat("Genes retained:", nrow(expr_filtered), "\n")

# --- Step 2: IQR-based per-gene winsorizing ---
# Caps (does not delete) values outside Q1 - 1.5*IQR / Q3 + 1.5*IQR
winsorize_row <- function(x) {
  q1  <- quantile(x, 0.25, na.rm = TRUE)
  q3  <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  x[x < (q1 - 1.5 * iqr)] <- q1 - 1.5 * iqr
  x[x > (q3 + 1.5 * iqr)] <- q3 + 1.5 * iqr
  x
}
expr_clean <- t(apply(expr_filtered, 1, winsorize_row))
colnames(expr_clean) <- colnames(expr_filtered)

# --- Step 3: Flag library-size outlier samples (|z| > 3) ---
lib_totals      <- colSums(expr_filtered)
z_lib           <- (lib_totals - mean(lib_totals)) / sd(lib_totals)
flagged_samples <- names(z_lib)[abs(z_lib) > 3]
cat("Samples flagged as library-size outliers (|z| > 3):",
    if (length(flagged_samples) > 0)
      paste(flagged_samples, collapse = ", ") else "none", "\n")
```

```{r pipeline-waterfall, fig.width=6, fig.height=4}
# Figure 2: Gene count at each cleaning step
steps_df <- data.frame(
  Step  = factor(c("Raw input", "After low-expr\nfiltering",
                    "After missing-\nvalue step"),
                  levels = c("Raw input", "After low-expr\nfiltering",
                              "After missing-\nvalue step")),
  Genes = c(nrow(expr_mat), nrow(expr_filtered), nrow(expr_filtered))
)
ggplot(steps_df, aes(x = Step, y = Genes)) +
  geom_col(fill = c("#B8860B", "#2E7D6E", "#2E7D6E"), width = 0.5) +
  geom_text(aes(label = scales::comma(Genes)), vjust = -0.4,
            fontface = "bold") +
  theme_minimal() +
  labs(title = "Gene Count at Each Cleaning Step",
       x = NULL, y = "Number of Genes") +
  scale_y_continuous(labels = scales::comma,
                     limits = c(0, nrow(expr_mat) * 1.12))
```

*Figure 2. Gene count at each stage of the cleaning pipeline. The
missing-value step removed no additional genes (0 missing cells).*

# 4. Missing Value Handling

```{r missing-handling}
# Check missingness
na_per_gene <- rowSums(is.na(expr_clean))
prop_na     <- na_per_gene / ncol(expr_clean)
total_na    <- sum(is.na(expr_clean))
cat("Total missing cells:", total_na, "\n")
cat("Percentage missing:", round(100 * total_na / length(expr_clean), 3), "%\n")

# Decision rule (per session material):
# > 20 % missing  -> remove gene ("ignore the tuple")
# <= 20 % missing -> impute with row median ("fill with most probable value")
high_na_genes <- prop_na > 0.20
cat("Genes removed for > 20% missingness:", sum(high_na_genes), "\n")
expr_complete <- expr_clean[!high_na_genes, ]

impute_median <- function(x) { x[is.na(x)] <- median(x, na.rm = TRUE); x }
expr_complete <- t(apply(expr_complete, 1, impute_median))
colnames(expr_complete) <- colnames(expr_clean)

cat("Any remaining missing values:", any(is.na(expr_complete)), "\n")
cat("Final matrix:", nrow(expr_complete), "genes x",
    ncol(expr_complete), "samples\n")
```

TCGA RSEM output assigns a zero count rather than `NA` for undetected
genes, so structural zeros (genuine biological silence) vastly
outnumber true missingness. As confirmed above, no genes required
removal and no imputation was performed — the decision framework was
applied, but the data were already complete.

# 5. Descriptive Statistics

Central tendency (mean, median), dispersion (standard deviation, IQR,
range), and relative variability (coefficient of variation, CV =
SD / mean × 100) were computed per gene, in the style of the
*Statistical Summary of Gene Expression* framework from the Basic
Statistics session.

```{r descriptive-stats}
gene_stats <- data.frame(
  gene   = rownames(expr_complete),
  mean   = rowMeans(expr_complete),
  median = apply(expr_complete, 1, median),
  sd     = apply(expr_complete, 1, sd),
  min    = apply(expr_complete, 1, min),
  max    = apply(expr_complete, 1, max),
  IQR    = apply(expr_complete, 1, IQR)
) %>%
  mutate(CV = round((sd / mean) * 100, 1)) %>%
  arrange(desc(sd))

# --- Dataset-level overall summary ---
overall_summary <- data.frame(
  Statistic = c("Genes (after cleaning)", "Samples",
                "Mean expression (raw counts)",
                "Median expression (raw counts)",
                "SD expression", "Min", "Max"),
  Value = c(nrow(expr_complete),
            ncol(expr_complete),
            round(mean(expr_complete), 2),
            round(median(expr_complete), 2),
            round(sd(as.vector(expr_complete)), 2),
            round(min(expr_complete), 2),
            round(max(expr_complete), 2))
)
overall_summary

# --- Top 15 most variable genes ---
head(gene_stats, 15)
```

The large gap between the mean (`r round(mean(expr_complete), 0)`) and
median (`r round(median(expr_complete), 0)`) raw counts confirms the
strong right-skew typical of RNA-seq count data, motivating the log
transformation in Section 7. The CV ranks genes by variability relative
to their expression level and is used in Section 8 for feature selection.

# 6. Data Visualization

```{r viz-sd-bar, fig.width=8, fig.height=5}
# Figure 3: Top 15 genes by SD
top15_plot <- head(gene_stats, 15) %>%
  mutate(gene = factor(gene, levels = rev(gene)))

ggplot(top15_plot, aes(x = gene, y = sd / 1e3)) +
  geom_col(fill = "#2E7D6E", alpha = 0.85) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 15 Most Variable Genes by Standard Deviation",
       x = NULL, y = "Standard Deviation (×1,000 raw counts)")
```

*Figure 3. Standard deviation of the 15 most variable genes in the
cleaned dataset.*

```{r viz-cv-bar, fig.width=8, fig.height=4}
# Figure 4: CV for top 15 genes, color-coded by threshold
cv15 <- head(gene_stats, 15) %>%
  mutate(
    gene    = factor(gene, levels = rev(gene_stats$gene[1:15])),
    cv_band = case_when(CV > 100 ~ "CV > 100%",
                        CV > 70  ~ "70% < CV ≤ 100%",
                        TRUE     ~ "CV ≤ 70%")
  )
ggplot(cv15, aes(x = gene, y = CV, fill = cv_band)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("CV > 100%"        = "#B22222",
                                "70% < CV ≤ 100%" = "#B8860B",
                                "CV ≤ 70%"        = "#2E7D6E"),
                    name = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Coefficient of Variation for Top 15 Variable Genes",
       x = NULL, y = "CV (%)")
```

*Figure 4. Coefficient of variation (CV) for the top 15 most variable
genes. Red bars (CV > 100%) denote genes with expression variability
exceeding their mean expression level.*

```{r viz-boxplot, fig.width=9, fig.height=5}
# Figure 5: Per-sample distribution (first 25 samples)
n_box   <- min(25, ncol(expr_complete))
long_df <- melt(expr_complete[, 1:n_box])
colnames(long_df) <- c("gene", "sample", "count")

ggplot(long_df, aes(x = sample, y = log2(count + 1))) +
  geom_boxplot(fill = "#2E7D6E", outlier.size = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Per-sample expression distribution (first 25 samples)",
       x = "Sample", y = "log2(raw count + 1)")
```

*Figure 5. Log2 expression distribution per sample. Spread and median
shifts flag samples worth closer review.*

```{r viz-histogram, fig.width=7, fig.height=4}
# Figure 6: Overall expression distribution (raw, 99th-percentile trimmed)
ggplot(data.frame(value = as.vector(expr_complete)), aes(x = value)) +
  geom_histogram(bins = 60, fill = "#B8860B") +
  xlim(0, quantile(expr_complete, 0.99)) +
  theme_minimal() +
  labs(title = "Overall expression distribution (99th percentile trimmed)",
       x = "Raw count", y = "Frequency")
```

*Figure 6. Strong right-skew in raw count distribution across all genes
and samples, confirming the need for log transformation.*

```{r viz-heatmap, fig.width=8, fig.height=6}
# Figure 7: Heatmap of top 50 most variable genes
top50 <- head(gene_stats$gene, 50)
pheatmap(log2(expr_complete[top50, ] + 1),
         show_rownames = FALSE, show_colnames = FALSE,
         main = "Heatmap: top 50 most variable genes (log2 scale)")
```

*Figure 7. Heatmap of the 50 most variable genes (log2 scale), giving
an early visual read on sample structure before formal PCA.*

# 7. Data Transformation

Two transformations are applied sequentially, both from the
Normalization Methods material.

```{r transform}
# log2(x + 1): compresses the large raw-count dynamic range
expr_log    <- log2(expr_complete + 1)

# Per-gene z-score: mean-centres and scales each gene to unit variance
# (necessary before PCA so that highly expressed genes do not dominate
# principal components regardless of biological informativeness)
expr_scaled <- t(scale(t(expr_log)))

cat("Post-scaling per-gene mean (should be ~0):",
    round(mean(rowMeans(expr_scaled, na.rm = TRUE)), 4), "\n")
cat("Post-scaling per-gene SD  (should be ~1):",
    round(mean(apply(expr_scaled, 1, sd, na.rm = TRUE)), 4), "\n")
```

# 8. Dimensionality Reduction: Feature Selection

```{r feature-selection}
# Select top 2,000 genes by CV (most variable relative to mean expression)
n_top      <- 2000
top_genes  <- gene_stats %>%
  arrange(desc(CV)) %>%
  slice_head(n = n_top) %>%
  pull(gene)

expr_reduced <- expr_scaled[top_genes, ]
cat("Reduced matrix for PCA:", nrow(expr_reduced), "genes x",
    ncol(expr_reduced), "samples\n")
```

Feature selection by CV reduces the `r nrow(expr_complete)` genes to
`r n_top`, removing low-variance genes that contribute noise to the
covariance structure PCA relies on.

# 9. Principal Component Analysis (PCA)

```{r pca}
pca_res      <- prcomp(t(expr_reduced), center = FALSE, scale. = FALSE)
var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

pca_variance_table <- data.frame(
  PC             = paste0("PC", 1:10),
  Variance_pct   = round(var_explained[1:10], 2),
  Cumulative_pct = round(cumsum(var_explained[1:10]), 2)
)
pca_variance_table
```

```{r pca-scree, fig.width=7, fig.height=4}
# Figure 8: Scree plot with cumulative variance line
scree_df <- data.frame(
  PC         = 1:10,
  Variance   = var_explained[1:10],
  Cumulative = cumsum(var_explained[1:10])
)

ggplot(scree_df, aes(x = PC)) +
  geom_col(aes(y = Variance), fill = "#2E7D6E", alpha = 0.9) +
  geom_line(aes(y = Cumulative), color = "#B8860B",
            linewidth = 1.2, group = 1) +
  geom_point(aes(y = Cumulative), color = "#B8860B", size = 2.5) +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal() +
  labs(title = "Scree Plot: Variance Explained by First 10 PCs",
       x = "Principal Component",
       y = "% Variance Explained  |  Cumulative %")
```

*Figure 8. Bars = individual variance explained; gold line = cumulative
variance across the first 10 PCs.*

```{r pca-scatter, fig.width=8, fig.height=6}
# Figure 9: PCA scatter colored by sample group (Tumor / Normal / Metastatic)
pca_df <- data.frame(
  PC1   = pca_res$x[, 1],
  PC2   = pca_res$x[, 2],
  Group = sample_group
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(aes(size = Group, alpha = Group)) +
  scale_color_manual(values = c("Tumor"      = "#B22222",
                                 "Normal"     = "#2E7D6E",
                                 "Metastatic" = "#6A0DAD")) +
  scale_size_manual(values  = c("Tumor" = 1.5, "Normal" = 2, "Metastatic" = 4)) +
  scale_alpha_manual(values = c("Tumor" = 0.55, "Normal" = 0.85,
                                 "Metastatic" = 1.0)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  theme_minimal() +
  labs(
    title = "PCA of TCGA-BRCA RNA-Seq Data",
    subtitle = paste0("Top ", n_top, " most variable genes  |  ",
                      "log2 + z-score transformed"),
    x     = paste0("PC1 (", round(var_explained[1], 1), "% variance explained)"),
    y     = paste0("PC2 (", round(var_explained[2], 1), "% variance explained)"),
    color = "Group", size = "Group", alpha = "Group"
  )
```

*Figure 9. PCA biplot of all 878 TCGA-BRCA samples. Red = Tumor
(n=775), teal = Normal (n=100), purple = Metastatic (n=3).*

PC1 (`r round(var_explained[1], 1)`%) and PC2
(`r round(var_explained[2], 1)`%) together explain
`r round(var_explained[1] + var_explained[2], 1)`% of the total
variance. PC1 achieves clear separation between tumor and normal
samples, confirming that the dominant axis of transcriptomic variation
corresponds to cancer status rather than technical or batch effects.
Normal samples cluster tightly on one side of PC1; tumor samples span
a broader range, reflecting the known molecular heterogeneity of breast
cancer subtypes (Luminal A/B, HER2-enriched, Basal-like). The three
metastatic samples fall within the tumor cloud, consistent with their
shared cancer-tissue biology.

# 10. Sample Proximity and Distance Measures

```{r proximity}
sample_dist <- as.matrix(dist(t(expr_reduced), method = "euclidean"))
sample_corr <- cor(expr_reduced, method = "pearson")
```

```{r proximity-euclidean, fig.width=7, fig.height=6}
# Figure 10: Euclidean distance heatmap
pheatmap(sample_dist,
         show_rownames = FALSE, show_colnames = FALSE,
         main = "Euclidean Distance Between Samples")
```

*Figure 10. Euclidean distance heatmap. Darker cells = smaller
distance (more similar expression profiles).*

```{r proximity-pearson, fig.width=7, fig.height=6}
# Figure 11: Pearson correlation heatmap
pheatmap(sample_corr,
         show_rownames = FALSE, show_colnames = FALSE,
         color = colorRampPalette(c("#2E7D6E", "white", "#B22222"))(50),
         main = "Pearson Correlation Between Samples")
```

*Figure 11. Pearson correlation heatmap. Red = high positive
correlation (similar expression profiles); teal = low/negative.*

Euclidean distance and Pearson correlation are the two continuous
proximity measures applicable here, as covered in the Proximity
Measures session. Categorical proximity measures (Simple Matching
Coefficient, Jaccard, chi-square) are not applicable because every
variable in the expression matrix is continuous numeric; they would
become relevant if categorical clinical metadata (tumor stage, receptor
status, PAM50 subtype) were joined to this dataset.

# 11. Summary of Methodology

| Step | Method | Rationale |
|---|---|---|
| Data Structuring | Barcode parsing; gene ID splitting; unmapped gene removal; duplicate symbol collapsing | TCGA text export stores 3 value types per sample; encodes IDs as `symbol\|entrez`; can carry unusable or duplicate identifiers |
| Noise Removal | CPM low-expression filter (1 CPM, ≥20% samples); IQR winsorizing (Q1±1.5×IQR); library-size z-score flag | Removes sequencing noise at gene and sample level without discarding observations |
| Missing Values | Check only — no imputation required (0 missing cells) | TCGA RSEM assigns zero, not NA, for undetected genes |
| Descriptive Statistics | Mean, median, SD, IQR, min, max, CV per gene; dataset-level summary | Covers central tendency, dispersion, and relative variability per session framework |
| Visualization | Sample composition bar; pipeline waterfall; SD/CV bars; per-sample boxplot; expression histogram; heatmap | Characterizes sample structure, gene variability, distribution shape, and multivariate relationships |
| Transformation | log2(x+1); per-gene z-score standardization | Compresses right-skew; equalizes gene variance for PCA |
| Dimensionality Reduction | CV-ranked feature selection (top 2,000 genes) | Removes low-signal genes before PCA to reduce noise in covariance structure |
| PCA | `prcomp()` on log2 + z-score matrix of top 2,000 genes | Summarizes dominant variance; PC1 separates tumor from normal |
| Proximity Measures | Euclidean distance + Pearson correlation between samples | Formalizes sample similarity in the reduced feature space; cross-checks PCA structure |

# 12. Conclusion

This workflow applied a complete data-mining pipeline to the TCGA-BRCA
RNA-Seq dataset (`r nrow(expr_mat)` initial genes, `r ncol(expr_mat)`
samples). After structuring, low-expression gene filtering (removing
`r sum(!keep_genes)` genes), and log2 + z-score transformation, PCA
of the top `r n_top` most variable genes revealed that the dominant
axis of transcriptomic variation (PC1,
`r round(var_explained[1], 1)`% of variance) cleanly separates primary
tumor from normal tissue — validating that the cleaning and
transformation steps preserved genuine biological signal rather than
introducing or amplifying technical noise.

Within-tumor spread along PC2 reflects the known molecular
heterogeneity of breast cancer subtypes. The three metastatic samples
cluster within the tumor group, consistent with shared cancer-tissue
biology. The Euclidean distance and Pearson correlation heatmaps
corroborate the PCA grouping. These findings provide a solid foundation
for downstream analyses such as consensus molecular subtyping,
differential expression, or survival modelling.

Every method applied was drawn directly from the taught material: IQR
outlier bounds, the missing-data decision framework, CV-based
variability ranking, log transformation, PCA as the capstone
dimensionality-reduction technique, and Euclidean/Pearson proximity
measures.
